SIR-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-03-20 paper_draft/main.tex
small fixes

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define two cost functions \cref{eq:cost_SIR,eq:cost_I} to calibrate the model with the use of huber loss~\cite{huberloss}. Our model typically uses \cref{eq:cost_SIR}, considering all three compartments: susceptible, infected, and recovered. However, there are instances where we need to balance modeling all groups and focusing on the infected group, which largely drives disease spread. In such cases, we consider losses from both \cref{eq:cost_SIR,eq:cost_I}. A weighted sum of the loss functions allows for trade-offs between comprehensive modeling and focusing on infected group dynamics (see \cref{optimizing_window_length}, where a hyperparameter(window length) is selected keeping both these losses in mind).
\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.} \\
  \end{cases}
  \label{eq:huberloss}
\end{equation}
In the above equation, $y$ is the actual data and $f(x)$ is the prediction.
\begin{equation}
  \begin{split}
  \textrm{loss\_SIR} = \textrm{cost\_function\_SIR}(S, \hat{S}, I, \hat{I}, R, \hat{R}) \\
  = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \end{split}
  \label{eq:cost_SIR}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = \textrm{cost\_function\_I}(I, \hat{I}) = L_{\delta = 1}(I, \hat{I})
  \label{eq:cost_I}
\end{equation}
Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Using the equations \Crefrange{eq:S_without_lockdown}{eq:R_without_lockdown}, we minimize the cost function \cref{eq:cost_SIR_without_lockdown} using the Nelder-Mead method~\cite{Gao2010} to estimate the parameters like $\beta$, $\gamma$, and fit the model to actual data. The following parameters and loss is obtained:
\begin{equation}
  \beta_{optimal} = 0.042
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.024
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.762
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 85051490.533
  \label{eq:cost_SIR_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 45187665.281
  \label{eq:cost_I_without_lockdown}
\end{equation}
See \cref{fig:SIR_model_IND} to see how the model compares with the actual data.

@@

\subsection{Optimizing Window Length for Time-varying Vaccination Rate}\label{optimizing_window_length}
However, as observed by the value of $\nu_{optimal}$ from \cref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible and the overestimation of infected individuals in \cref{fig:SIR_model_with_lockdown_with_vaccination_infections_IND} suggests that $\nu$ might be varying with time. This suggests to accurately estimate the infected population a time-varying vaccination rate should be used as the transition from susceptibility to direct recovery fluctuates with time~\cite{Liang2021,Marinov2022}. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \cref{eq:beta_optimal_with_lockdown_and_nu} and \cref{eq:gamma_optimal_with_lockdown_and_nu}, we first find the optimal window length~\cite{Liao2020} i.e., a window length that results in the least loss from \cref{eq:cost_SIR,eq:cost_I}. Using different window lengths ($\textrm{window\_length} = 5, 10, 15 \dots 40, 45, 50$), we estimate value of $\nu$ given a sub-interval of $[{start}, {start} + {window\_length}]$ then for the next time-step $[{start} + {window\_length}, {start} + {window\_length} + {window\_length}]$ we again estimate $\nu$ -- so on and so forth. It is crucial to note that the variable $\nu$ is constrained to be a positive integer, reflecting the inherent one-way nature of vaccination: individuals can only receive vaccinations, not return them.
\begin{equation}
  \frac{dS}{dt} = -\beta_{optimal}  (1 - s(t)/100)  \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta_{optimal}  (1 - s(t)/100) \frac{S I}{N} - \gamma_{optimal} I
  \label{eq:I_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma_{optimal} I + \nu S
  \label{eq:R_with_lockdown_and_nu_calc_nu}
\end{equation}
This model gives shows how $\nu$ varies with time. Since, the window length acts like a hyperparameter, we can try different window lengths in order to evaluate which would give us the best results. Therefore, the below graphs show us the loss using \cref{eq:cost_SIR,eq:cost_I}.

@@

In this case of epidemic control by non-pharmaceutical interventions (NPI) based strategies this policy represents the series of stringencies to be imposed upon the population to shift the initial status of the environment to a targeted status which is equivalent to the desired set of system states. This is how the $Q$-table updates saying, if in state $s_k$ the most ideal action is $a_t$. After having more and more experience with the environment and understanding which actions lead to a higher reward $r$ an optimal policy is derived be maximizing the expected value of discounted reward $J\left(r_t\right)=\mathbb{E}\left[\sum_{t=1}^{\infty} \gamma^{(t-1)} r_t\right]$, where, discount factor $\gamma \in [0, 1]$ and time steps $k = 1, 2, \dots$.

@@

Additionally, if the proportion of the infected population were to rise above 0.003(peak in the actual data) the model is punished ($-2000$) and otherwise rewarded ($50$). To reward not changing the stringencies frequently, we reward the absolute different between the previous stringency and the current stringency negatively ($|s(t) - s(t-1)| \times -12$).

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=0.7\linewidth]{images/sir_arch.pdf}
  \caption{\textbf{Deep Reinforcement Learning.} An integration of deep learning and reinforcement learning to enable more complex decision-making.}
  \label{fig:sir_arch}
\end{figure}

@@

In contrast, an alternative output is presented in \cref{fig:175975_parent}, which demonstrates a gradual increase in stringency from October, 2020 till April, 2021. This strategy results in a decline of infections occurring prior to the vaccine's release, and a subsequent cessation of new infections. This approach, however, has implications for the normalized GDP, as seen by the observed decline in \cref{fig:175975_rl_gdp}. These findings underscore the complexity of managing public health crises and the need for careful strategic planning to balance health outcomes with economic considerations.

>>>>

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define two cost functions \cref{eq:cost_SIR,eq:cost_I} to calibrate the model with the use of huber loss~\cite{huberloss}. Our model typically uses \cref{eq:cost_SIR}, considering all three compartments: susceptible, infected, and recovered. However, there are instances where we need to balance modeling all groups and focusing on the infected group (population that drives disease spread). In such cases, we consider losses from both \cref{eq:cost_SIR,eq:cost_I}. A weighted sum of the loss functions allows for trade-offs between comprehensive modeling and focusing on infected group dynamics (see \cref{optimizing_window_length}, where a hyperparameter (window length) is selected keeping both these losses in mind).
\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.} \\
  \end{cases}
  \label{eq:huberloss}
\end{equation}
In the above equation, $y$ is the actual data and $f(x)$ is the prediction.
\begin{equation}
  \begin{split}
  \textrm{loss\_SIR} = \textrm{cost\_function\_SIR}(S, \hat{S}, I, \hat{I}, R, \hat{R}) \\
  = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \end{split}
  \label{eq:cost_SIR}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = \textrm{cost\_function\_I}(I, \hat{I}) = L_{\delta = 1}(I, \hat{I})
  \label{eq:cost_I}
\end{equation}
Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Using the equations \Crefrange{eq:S_without_lockdown}{eq:R_without_lockdown}, we minimize the cost function \cref{eq:cost_SIR_without_lockdown} using the Nelder-Mead method~\cite{Gao2010} to estimate the parameters like $\beta$, $\gamma$, and fit the model to actual data. The following parameters and loss is obtained:
\begin{equation}
  \beta_{optimal} = 0.042
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.024
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.762
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 85051490.533
  \label{eq:cost_SIR_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 45187665.281
  \label{eq:cost_I_without_lockdown}
\end{equation}
See \cref{fig:SIR_model_IND} to see how the model compares with the actual data. [TODO: should we add this to all the models or remove it for the first mention and let the reader read in the results' section which equations were used??]

@@

\subsection{Optimizing Window Length for Time-varying Vaccination Rate}\label{optimizing_window_length}
However, as observed by the value of $\nu_{optimal}$ from \cref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible and the overestimation of infected individuals in \cref{fig:SIR_model_with_lockdown_with_vaccination_infections_IND} suggests that $\nu$ might be varying with time. This suggests to accurately estimate the infected population a time-varying vaccination rate should be used as the transition from susceptibility to direct recovery fluctuates with time~\cite{Liang2021,Marinov2022}. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \cref{eq:beta_optimal_with_lockdown_and_nu} and \cref{eq:gamma_optimal_with_lockdown_and_nu}, we first find the optimal window length~\cite{Liao2020} for which the value of $\nu$ is constant and results in the least loss from \cref{eq:cost_SIR,eq:cost_I}. Using different window lengths ($\textrm{window\_length} = 5, 10, 15 \dots 40, 45, 50$), we estimate value of $\nu$ given a sub-interval of $[{start}, {start} + {window\_length}]$ then for the next time-step $[{start} + {window\_length}, {start} + {window\_length} + {window\_length}]$ we again estimate $\nu$ -- so on and so forth. It is crucial to note that the variable $\nu$ is constrained to be a positive integer, reflecting the inherent one-way nature of vaccination: individuals can only receive vaccinations, not return them.
\begin{equation}
  \frac{dS}{dt} = -\beta_{optimal}  (1 - s(t)/100)  \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta_{optimal}  (1 - s(t)/100) \frac{S I}{N} - \gamma_{optimal} I
  \label{eq:I_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma_{optimal} I + \nu S
  \label{eq:R_with_lockdown_and_nu_calc_nu}
\end{equation}
% This model gives shows how $\nu$ varies with time. Since, the window length acts like a hyperparameter, we can try different window lengths in order to evaluate which would give us the best results. Therefore, the below graphs show us the loss using \cref{eq:cost_SIR,eq:cost_I}.

@@

In this case of epidemic control by non-pharmaceutical interventions (NPI) based strategies this policy represents the series of stringencies to be imposed upon the population to shift the initial status of the environment to a targeted status which is equivalent to the desired set of system states. This is how the $Q$-table updates saying, if in state $s_k$ the most ideal action is $a_t$. After having more and more experience with the environment and understanding which actions lead to a higher reward $r$ an optimal policy is derived be maximizing the expected value of discounted reward $J\left(r_t\right)=\mathbb{E}\left[\sum_{t=1}^{\infty} \gamma^{(t-1)} r_t\right]$, where, discount factor $\gamma \in [0, 1]$ (in our case $\gamma = 0.99$) and time steps $k = 1, 2, \dots$.

@@

Additionally, if the proportion of the infected population were to rise above 0.003 (peak in the actual data) the model is punished ($-2000$) and otherwise rewarded ($50$). To reward not changing the stringencies frequently, we reward the absolute different between the previous stringency and the current stringency negatively ($|s(t) - s(t-1)| \times -12$).

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=0.7\linewidth]{images/sir_arch.pdf}
  \caption{\textbf{Deep Reinforcement Learning.} Deep learning algorithms used in reinforcement learning enables more complex decision-making.}
  \label{fig:sir_arch}
\end{figure}

@@

In contrast, an alternative output is presented in \cref{fig:175975_parent}, which demonstrates a gradual increase in stringency from October, 2020 till April, 2021. This strategy results in a decline of infections occurring prior to the vaccine's release, and a subsequent cessation of new infections. This approach, however, has implications for the normalized GDP, as seen by the observed decline in \cref{fig:175975_rl_gdp}. While both these approaches (\cref{fig:175975_parent,fig:176647_parent}) outperform the actual strategy, they underscore the complexity of managing public health crises and the need for careful strategic planning to balance health outcomes with economic considerations.
